### 2023-08-23 documents/config/cariban_language_names.tex
more table fixes

@@

\providecommand{\kapon}{Kapón\xspace}
\DeclareAcronym{kap_lg}{short=kap,long=Kapón,short-format=\scshape}
\providecommand{\carijo}{Carijona\xspace}
\DeclareAcronym{car_lg}{short=car,long=Carijona,short-format=\scshape}

@@

@@

\providecommand{\sikiana}{†Sikiana\xspace}
\DeclareAcronym{siku_lg}{short=siku,long=†Sikiana,short-format=\scshape}

@@

\DeclareAcronym{chai_lg}{short=chai,long=†Chayma,short-format=\scshape}

@@

\DeclareAcronym{pemo_lg}{short=pemo,long=†Pémono,short-format=\scshape}

@@

\DeclareAcronym{pc_lg}{short=pc,long=Proto-Cariban,short-format=\scshape}
\newcommand{\PPar}{Proto-Parukotoan\xspace}
\newcommand{\PPars}{\textsc{PPar}\xspace}
\newcommand{\PPem}{Proto-Pemongan\xspace}
\newcommand{\PPems}{\textsc{PPem}\xspace}
\newcommand{\PPek}{Proto-Pekodian\xspace}
\newcommand{\PPeks}{\textsc{PPek}\xspace}
\newcommand{\PPP}{Proto-Panare-Pemongan\xspace}
\newcommand{\PPPs}{\textsc{PPP}\xspace}
\newcommand{\PTir}{Proto-Tiriyoan\xspace}
\newcommand{\PTirs}{\textsc{PTir}\xspace}
\newcommand{\PWai}{Proto-Waiwaian\xspace}
\newcommand{\PWais}{\textsc{PWai}\xspace}
\newcommand{\PXin}{Proto-Xinguan\xspace}
\newcommand{\PXins}{\textsc{PXin}\xspace}
\newcommand{\PTar}{Proto-Taranoan\xspace}
\newcommand{\PTars}{\textsc{PTar}\xspace}
\newcommand{\yabmap}{\mapoyo-\yawarana}
\newcommand{\yabmaps}{\textsc{yab-map}\xspace}
\newcommand{\vzk}{Venezuelan \kalina}
\newcommand{\vzks}{\textsc{vzk}\xspace}
\newcommand{\esk}{Eastern Surinamese \kalina}
\newcommand{\esks}{\textsc{esk}\xspace}
\newcommand{\wsk}{Western Surinamese \kalina}
\newcommand{\wsks}{\textsc{wsk}\xspace}
\renewcommand{\lg}[1]{\acs*{#1_lg}}

>>>>

\providecommand{\wbakairi}{Western Bakairi\xspace}
\DeclareAcronym{wbak_lg}{short=wbak,long=Western Bakairi,short-format=\scshape}
\providecommand{\ebakairi}{Eastern Bakairi\xspace}
\DeclareAcronym{ebak_lg}{short=ebak,long=Eastern Bakairi,short-format=\scshape}
\providecommand{\tabajari}{Tabajari\xspace}
\DeclareAcronym{tab_lg}{short=tab,long=Tabajari,short-format=\scshape}
\providecommand{\gkar}{Guianese Carib\xspace}
\DeclareAcronym{gkar_lg}{short=gkar,long=Guianese Carib,short-format=\scshape}
\providecommand{\aretyry}{Aretyry\xspace}
\DeclareAcronym{aty_lg}{short=aty,long=Aretyry,short-format=\scshape}
\providecommand{\sikiana}{Sikiana\xspace}
\DeclareAcronym{siku_lg}{short=siku,long=Sikiana,short-format=\scshape}
\providecommand{\nahukwa}{Nahukwa\xspace}
\DeclareAcronym{nah_lg}{short=nah,long=Nahukwa,short-format=\scshape}
\providecommand{\matipu}{Matipu\xspace}
\DeclareAcronym{mat_lg}{short=mat,long=Matipu,short-format=\scshape}
\providecommand{\carijo}{Karijona\xspace}
\DeclareAcronym{car_lg}{short=car,long=Karijona,short-format=\scshape}
\providecommand{\kapon}{Kapong\xspace}
\DeclareAcronym{kap_lg}{short=kap,long=Kapong,short-format=\scshape}

@@

@@

\DeclareAcronym{cha_lg}{short=cha,long=†Chayma,short-format=\scshape}

@@

@@

\DeclareAcronym{pno_lg}{short=pno,long=†Pémono,short-format=\scshape}

@@

@@

\DeclareAcronym{PC_lg}{short=PC,long=Proto-Cariban,short-format=\scshape}
\providecommand{\PPar}{Proto-Parukotoan\xspace}
\DeclareAcronym{PPar_lg}{short=PPar,long=Proto-Parukotoan,short-format=\scshape}
\providecommand{\PWai}{Proto-Waiwaian\xspace}
\DeclareAcronym{PWai_lg}{short=PWai,long=Proto-Waiwaian,short-format=\scshape}
\providecommand{\PPek}{Proto-Pekodian\xspace}
\DeclareAcronym{PPek_lg}{short=PPek,long=Proto-Pekodian,short-format=\scshape}
\providecommand{\PXin}{Proto-Xinguan\xspace}
\DeclareAcronym{PXin_lg}{short=PXin,long=Proto-Xinguan,short-format=\scshape}
\providecommand{\PPP}{Proto-Pemong-Panare\xspace}
\DeclareAcronym{PPP_lg}{short=PPP,long=Proto-Pemong-Panare,short-format=\scshape}
\providecommand{\PPem}{Proto-Pemongan\xspace}
\DeclareAcronym{PPem_lg}{short=PPem,long=Proto-Pemongan,short-format=\scshape}
\providecommand{\PTar}{Proto-Taranoan\xspace}
\DeclareAcronym{PTar_lg}{short=PTar,long=Proto-Taranoan,short-format=\scshape}
\providecommand{\PTir}{Proto-Tiriyoan\xspace}
\DeclareAcronym{PTir_lg}{short=PTir,long=Proto-Tiriyoan,short-format=\scshape}
\providecommand{\PMan}{Proto-Manapiarean\xspace}
\DeclareAcronym{PMan_lg}{short=PMan,long=Proto-Manapiarean,short-format=\scshape}
\providecommand{\PVen}{Proto-Venezuelan\xspace}
\DeclareAcronym{PVen_lg}{short=PVen,long=Proto-Venezuelan,short-format=\scshape}
\providecommand{\PCum}{Proto-Cumana\xspace}
\DeclareAcronym{PCum_lg}{short=PCum,long=Proto-Cumana,short-format=\scshape}

### 2023-08-23 documents/floats/comepara.tex
more table fixes

\begin{table}
\centering
\caption[\rc{(ət-)epɨ} \qu{to come} in paradigms]{\rc{(ət-)epɨ} \qu{to come} in paradigms \parencites[113, 150, 153, 156]{alves2017arara}[294; p.c., Spike Gildea]{triomeira1999}}
\label{tab:comepara}
\begin{tabular}[t]{@{}llll@{}}
\mytoprule
{} &          \kaxui &          \arara &          \trio \\
\mymidrule
\gl{1}   &   \obj{w-oohɨ-} &    \obj{w-ebɨ-} &  \obj{w-əepɨ-} \\
\gl{2}   &   \obj{m-oohɨ-} &  \obj{m-odebɨ-} &  \obj{mən-epɨ} \\
\gl{1+2} &  \obj{kɨs-ohɨ-} &  \obj{kud-ebɨ-} &   \obj{ke-epɨ} \\
\gl{3}   &    \obj{n-ehɨ-} &    \obj{t-ebɨ-} &    \obj{n-epɨ} \\
\mybottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption[\rc{(ət-)epɨ} \qu{to come} in paradigms]{\rc{(ət-)epɨ} \qu{to come} in paradigms \parencites[113, 150, 153, 156]{alves2017arara}[294]{triomeira1999}[Spike Gildea]{pc}}
\label{tab:comepara}
\begin{tabular}[t]{@{}llll@{}}
\mytoprule
{} &          \arara &          \trio &          \kaxui \\
\mymidrule
\gl{1}   &    \obj{w-ebɨ-} &  \obj{w-əepɨ-} &   \obj{w-oohɨ-} \\
\gl{2}   &  \obj{m-odebɨ-} &  \obj{mən-epɨ} &   \obj{m-oohɨ-} \\
\gl{1+2} &  \obj{kud-ebɨ-} &   \obj{ke-epɨ} &  \obj{kɨs-ohɨ-} \\
\gl{3}   &    \obj{t-ebɨ-} &    \obj{n-epɨ} &    \obj{n-ehɨ-} \\
\mybottomrule
\end{tabular}
\end{table}

### 2023-08-23 documents/floats/go.tex
more table fixes

\begin{table}
\centering
\caption[Reflexes of \rc{ɨtə[mə]} \qu{to go}]{Reflexes of \rc{ɨtə[mə]} \qu{to go} \parencites[291]{cruz2005fonologia}[292]{triomeira1999}[195]{wayanatavares2005}[87]{gildea1994akuriyo}[153]{alves2017arara}[27, 248]{hixkaryanaderby1985}[45, 62]{waiwaihawkins1998}[54, 80]{ikpengpacheco2001}[112, 374]{von1892bakairi}[181, 216]{maquiritaricaceres2011}[112]{meira1998proto}[168]{hoff1968carib}[139]{meira2006syntactic}[4]{caceres2018yawarana}[74]{muller1975mapoyo}[198]{mattei1994diccionario}[48, 50]{macushiabbott1991}[172]{garcia2006diccionario}[6]{franchetto2002kuikuro}[99; p.c., Spike Gildea]{camargo2002lexico}}
\label{tab:go}
\begin{tabular}[t]{@{}lllllll@{}}
\mytoprule
Language & \multicolumn{6}{l}{Form} \\
\midrule
\kaxui    &     \obj{to[mo]} &    &  t &  o &  m &  o \\
\hixka    &      \obj{[ɨ]to} &  ɨ &  t &  o &    &    \\
\waiwai   &   \obj{[e]to[m]} &  e &  t &  o &  m &    \\
\arara    &        \obj{ɨdo} &  ɨ &  d &  o &    &    \\
\ikpeng   &       \obj{aran} &  a &  r &  a &  n &    \\
\ikpeng   &        \obj{ero} &  e &  r &  o &    &    \\
\bakairi  &      \obj{[ɨ]tə} &  ɨ &  t &  ə &    &    \\
\trio     &      \obj{tə[n]} &    &  t &  ə &  n &    \\
\akuriyo  &  \obj{[ə]tə[mɨ]} &  ə &  t &  ə &  m &  ɨ \\
\carijo   &       \obj{təmə} &    &  t &  ə &  m &  ə \\
\wayana   &   \obj{[ɨ]tə[m]} &  ɨ &  t &  ə &  m &    \\
\apalai   &        \obj{ɨto} &  ɨ &  t &  o &    &    \\
\kalina   &         \obj{to} &    &  t &  o &    &    \\
\kalina   &       \obj{[ɨ]ʔ} &  ɨ &  ʔ &    &    &    \\
\maqui    &    \obj{ɨtə[mə]} &  ɨ &  t &  ə &  m &  ə \\
\ingariko &        \obj{ətə} &  ə &  t &  ə &    &    \\
\pemon    &      \obj{[e]tə} &  e &  t &  ə &    &    \\
\macushi  &      \obj{[a]tɨ} &  a &  t &  ɨ &    &    \\
\panare   &      \obj{tə[n]} &    &  t &  ə &  n &    \\
\yawarana &         \obj{tə} &    &  t &  ə &    &    \\
\mapoyo   &         \obj{tə} &    &  t &  ə &    &    \\
\uxc      &      \obj{[e]te} &  e &  t &  e &    &    \\
\yukpa    &         \obj{to} &    &  t &  o &    &    \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption[Reflexes of \rc{ɨtə[mə]} \qu{to go}]{Reflexes of \rc{ɨtə[mə]} \qu{to go} \parencites[291]{cruz2005fonologia}[292]{triomeira1999}[195]{wayanatavares2005}[87]{gildea1994akuriyo}[80, 153]{alves2017arara}[27, 248]{hixkaryanaderby1985}[45, 62]{waiwaihawkins1998}[54, 80]{ikpengpacheco2001}[112, 374]{von1892bakairi}[181, 216]{maquiritaricaceres2011}[112]{meira1998proto}[5]{guerrero2016karihona}[168]{hoff1968carib}[139]{meira2006syntactic}[4]{caceres2018yawarana}[74]{muller1975mapoyo}[198]{mattei1994diccionario}[48, 50]{macushiabbott1991}[172]{garcia2006diccionario}[Spike Gildea]{pc}[6]{franchetto2002kuikuro}[99]{camargo2002lexico}[14]{lacerda1991waimiri}}
\label{tab:go}
\begin{tabular}[t]{@{}ll@{}}
\mytoprule
Language &             Form \\
\mymidrule
\kaxui    &     \obj{to[mo]} \\
\PWai     &    \rc{[ɨ]to[m]} \\
\hixka    &      \obj{[ɨ]to} \\
\waiwai   &   \obj{[e]to[m]} \\
\PPek     &      \rc{ɨtə[n]} \\
\bakairi  &      \obj{[ɨ]tə} \\
\arara    &         \obj{to} \\
\arara    &        \obj{ɨdo} \\
\ikpeng   &       \obj{aran} \\
\ikpeng   &        \obj{ero} \\
\carijo   &     \obj{tə[mə]} \\
\PTir     &   \rc{[ə]tə[mɨ]} \\
\trio     &      \obj{tə[n]} \\
\akuriyo  &  \obj{[ə]tə[mɨ]} \\
\wayana   &   \obj{[ɨ]tə[m]} \\
\apalai   &        \obj{ɨto} \\
\kalina   &       \obj{[ɨ]ʔ} \\
\kalina   &         \obj{to} \\
\maqui    &    \obj{ɨtə[mə]} \\
\ingariko &        \obj{ətə} \\
\pemon    &      \obj{[e]tə} \\
\macushi  &      \obj{[a]tɨ} \\
\panare   &      \obj{tə[n]} \\
\yawarana &         \obj{tə} \\
\mapoyo   &         \obj{tə} \\
\uxc      &      \obj{[e]te} \\
\yukpa    &         \obj{to} \\
\waimiri  &         \obj{ɨʔ} \\
\mybottomrule
\end{tabular}
\end{table}

### 2023-08-23 documents/floats/pwaireg.tex
more table fixes

\begin{table}
\centering
\caption[Regular \PWai verbs]{Regular \qu{to fall} (\gl{s_a_}) and \qu{to sleep} (\gl{s_p_}) in \PWai \parencites[150]{howard2001wrought}[189-190, 196]{hixkaryanaderby1985}[30]{waiwaihawkins1998}[209, 211]{hawkins1953waiwai}}
\label{tab:pwaireg}
\begin{tabular}[t]{@{}lllllll@{}}
\mytoprule
{} &     \qu{to fall} &                  &                    &      \qu{to sleep} &                 &                    \\
{} &            \PWai &           \hixka &            \waiwai &              \PWai &          \hixka &            \waiwai \\
\mymidrule
\gl{1}   &  \obj{k-eɸurka-} &  \obj{k-ehurka-} &    \obj{k-eɸɨrka-} &   \obj{kɨ-wɨnɨkɨ-} &  \obj{kɨ-nɨkɨ-} &   \obj{kɨ-wɨnɨkɨ-} \\
\gl{2}   &  \obj{m-eɸurka-} &  \obj{m-ehurka-} &    \obj{m-eɸɨrka-} &    \obj{o-wɨnɨkɨ-} &  \obj{o-wnɨkɨ-} &   \obj{mɨ-wɨnɨkɨ-} \\
\gl{1+2} &  \obj{t-eɸurka-} &  \obj{t-ehurka-} &  \obj{t͡ʃ-eɸɨrka-} &  \obj{tɨt-wɨnɨkɨ-} &  \obj{tɨ-nɨkɨ-} &  \obj{tɨt-wɨnɨkɨ-} \\
\gl{3}   &  \obj{ɲ-eɸurka-} &  \obj{ɲ-ehurka-} &    \obj{ɲ-eɸɨrka-} &   \obj{nɨ-wɨnɨkɨ-} &  \obj{nɨ-nɨkɨ-} &   \obj{nɨ-wɨnɨkɨ-} \\
\mybottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption[Regular \PWai verbs]{Regular \qu{to fall} (\gl{s_a_}) and \qu{to sleep} (\gl{s_p_}) in \PWai \parencites[150]{howard2001wrought}[189, 196]{hixkaryanaderby1985}[30]{waiwaihawkins1998}[209, 211]{hawkins1953waiwai}}
\label{tab:pwaireg}
\begin{tabular}[t]{@{}lllllll@{}}
\mytoprule
{} &    \qu{to fall} &                  &                    &     \qu{to sleep} &                 &                    \\
{} &           \PWai &           \hixka &            \waiwai &             \PWai &          \hixka &            \waiwai \\
\mymidrule
\gl{1}   &  \rc{k-eɸurka-} &  \obj{k-ehurka-} &    \obj{k-eɸɨrka-} &   \rc{kɨ-wɨnɨkɨ-} &  \obj{kɨ-nɨkɨ-} &   \obj{kɨ-wɨnɨkɨ-} \\
\gl{2}   &  \rc{m-eɸurka-} &  \obj{m-ehurka-} &    \obj{m-eɸɨrka-} &    \rc{o-wɨnɨkɨ-} &  \obj{o-wnɨkɨ-} &   \obj{mɨ-wɨnɨkɨ-} \\
\gl{1+2} &  \rc{t-eɸurka-} &  \obj{t-ehurka-} &  \obj{t͡ʃ-eɸɨrka-} &  \rc{tɨt-wɨnɨkɨ-} &  \obj{tɨ-nɨkɨ-} &  \obj{tɨt-wɨnɨkɨ-} \\
\gl{3}   &  \rc{ɲ-eɸurka-} &  \obj{ɲ-ehurka-} &    \obj{ɲ-eɸɨrka-} &   \rc{nɨ-wɨnɨkɨ-} &  \obj{nɨ-nɨkɨ-} &   \obj{nɨ-wɨnɨkɨ-} \\
\mybottomrule
\end{tabular}
\end{table}

### 2023-08-23 documents/floats/say.tex
more table fixes

\begin{table}
\centering
\caption[Reflexes of \rc{ka[ti]} \qu{to say}]{Reflexes of \rc{ka[ti]} \qu{to say} \parencites[4]{meira2003bakairi}[48]{franchetto2008absolutivo}[209]{ikpengpacheco2001}[153]{alves2017arara}[182]{hixkaryanaderby1985}[113]{meira1998proto}[107]{koehn1986apalai}[26]{waiwaihawkins1998}[66]{camargo2010wayana}[59]{macushiabbott1991}[123]{swiggers2010gramatica}[430]{courtz2008carib}[125]{akawaiocaesar2003}[102]{mattei1994diccionario}[63; p.c., Spike Gildea]{largo2011yukpa}}
\label{tab:say}
\begin{tabular}[t]{@{}llllll@{}}
\mytoprule
Language & \multicolumn{5}{l}{Form} \\
\midrule
\kaxui   &   \obj{ka[s]} &  k &  a &  s &    \\
\hixka   &   \obj{ka[h]} &  k &  a &  h &    \\
\waiwai  &   \obj{ka[s]} &  k &  a &  s &    \\
\arara   &      \obj{ke} &  k &  e &    &    \\
\ikpeng  &      \obj{ke} &  k &  e &    &    \\
\bakairi &      \obj{ke} &  k &  e &    &    \\
\trio    &      \obj{ka} &  k &  a &    &    \\
\akuriyo &      \obj{ka} &  k &  a &    &    \\
\carijo  &      \obj{ka} &  k &  a &    &    \\
\wayana  &   \obj{ka[i]} &  k &  a &    &  i \\
\apalai  &  \obj{ka[ʃi]} &  k &  a &  ʃ &  i \\
\kalina  &      \obj{ka} &  k &  a &    &    \\
\kapon   &      \obj{ka} &  k &  a &    &    \\
\pemon   &      \obj{ka} &  k &  a &    &    \\
\macushi &      \obj{ka} &  k &  a &    &    \\
\panare  &   \obj{ka[h]} &  k &  a &  h &    \\
\uxc     &      \obj{ki} &  k &  i &    &    \\
\yukpa   &      \obj{ka} &  k &  a &    &    \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption[Reflexes of \rc{ka[ti]} \qu{to say}]{Reflexes of \rc{ka[ti]} \qu{to say} \parencites[267]{meira2005bakairi}[4]{meira2003bakairi}[48]{franchetto2008absolutivo}[209, 279]{ikpengpacheco2001}[80, 153]{alves2017arara}[182]{hixkaryanaderby1985}[113]{meira1998proto}[107]{koehn1986apalai}[26]{waiwaihawkins1998}[Spike Gildea]{pc}[66]{camargo2010wayana}[59]{macushiabbott1991}[123]{swiggers2010gramatica}[430]{courtz2008carib}[125]{akawaiocaesar2003}[102]{mattei1994diccionario}[63]{largo2011yukpa}}
\label{tab:say}
\begin{tabular}[t]{@{}ll@{}}
\mytoprule
Language &          Form \\
\mymidrule
\kaxui   &   \obj{ka[s]} \\
\PWai    &    \rc{ka[s]} \\
\hixka   &   \obj{ka[s]} \\
\waiwai  &   \obj{ka[s]} \\
\PPek    &       \rc{ge} \\
\PPek    &       \rc{ke} \\
\bakairi &      \obj{ge} \\
\bakairi &      \obj{ke} \\
\arara   &      \obj{ge} \\
\arara   &      \obj{ke} \\
\ikpeng  &      \obj{ge} \\
\ikpeng  &    \obj{k[e]} \\
\carijo  &      \obj{ka} \\
\PTir    &       \rc{ka} \\
\trio    &      \obj{ka} \\
\akuriyo &      \obj{ka} \\
\wayana  &   \obj{ka[i]} \\
\apalai  &  \obj{ka[ʃi]} \\
\kalina  &      \obj{ka} \\
\kapon   &      \obj{ka} \\
\pemon   &      \obj{ka} \\
\macushi &      \obj{ka} \\
\panare  &   \obj{ka[h]} \\
\uxc     &      \obj{ki} \\
\yukpa   &      \obj{ka} \\
\mybottomrule
\end{tabular}
\end{table}

### 2023-08-23 documents/sections/verbs.tex
more table fixes

As mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes \hixka \obj{ka[s]} as transitive, a choice not only motivated by a desire to avoid an idiosyncratic intransitive first person prefix \obj{ɨ-} instead of regular \obj{kɨ-}.
The verb also shows the complementary distribution of third person \obj{n-} and preceding objects typical of transitive verbs in Cariban \parencite[60--81]{gildea1998}.
Due to its semantics, these objects are either ideophones or direct speech \exref{hixsay}.

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}), while does not occur in \exref{hixsay.ordpl-35} where the verb is preceded by direct speech.
In the rest of the family, at least the \trio reflex shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

@@

\pex<sayintr>
\a<aku-154> \akuriyo \parencite[][113]{meira1998proto}\\
\begingl
\gla mɨ-ka//
\glb \gl{2}-say//
\glft \qu{You said.}//
\endgl
\a<kar-84> \kalina \parencite[][202]{courtz2008carib}\\
\begingl
\glpreamble Òmakon `wa oti ywykàpo kaiko.//
\gla o-ʔma-kon ʔwa oti ɨ-wɨ-ka-ʔpo kai-ko//
\glb \gl{2}-child-\gl{pl} \gl{obl} greeting \gl{1}-\gl{s_a_}-say-\gl{pst}.\gl{nmlz} say-\gl{imp}//
\glft \qu{Pass my greetings to your children.}//
\endgl
\xe

@@

\ex<mivida-12>\carijo \parencite[][177]{robayo1989rame}\\
\begingl
\glpreamble iretibə et͡ʃinəme gərə jet͡ʃiɨ//
\gla ireti-bə et͡ʃi-nə=me gərə j-et͡ʃi-ɨ//
\glb then-from be-\gl{inf}=\gl{attrz} still \gl{1}-be-\gl{pfv}//
\glft \qu{Then I was already grown up.}//
\endgl
\xe

@@

\ex<car-18>\carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\xe
%
\posscite{gildea2007greenberg} reconstruction \rc{ətepɨ} can be segmented into an optional prefix \rc{ət-} and a root \rc{epɨ}, since reflexes can be grouped into those with a reflex of \rc{ət-} and those without \pcref{tab:come}.
As seen in \cref{tab:comepara}, this division can exist within a single paradigm.

@@

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a family-wide perspective reveals a more complicated story \pcref{tab:godown}.%
\footnote{
The cognacy status of parenthesized forms in \cref{tab:godown} is uncertain.
The reconstruction of \PPek \rc{ɨptə} treats the additional elements in daughter languages as non-cognate.
\textcite{meira2005southern} identify no correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}, and at least the addition of a final \rc{ŋ} in \PXin is attested elsewhere:
		\begin{inlinelist}
			\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
			\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
			\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
		\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
While a verb \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about its class.
Its reflexes in languages that preserve the split-\gl{s} system are distributed fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

The verb shows traits of both classes in \wayana, making it a \dbqu{mixed} verb synchronically.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\begin{multicols}{2}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\end{multicols}
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the \PC  causativizer \rc{-ka} was restricted to \gl{s_p_} verbs \parencite{gildea2019overview}.
These patterns point to \qu{to go down} being a regular \gl{s_p_} verb in pre-\wayana, but partially switching to the \gl{s_a_} class by taking a \gl{1+2}\gl{s_a_} prefix and the \gl{s_a_} class marker.
This in turn implies that (inflectionally defined) \gl{s_a_} reflexes in other languages fully switched from \gl{s_p_}.

>>>>

As mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes \hixka \obj{ka[s]} as transitive, a choice not only motivated by a desire to avoid an idiosyncratic intransitive first person prefix \obj{ɨ-} instead of regular \obj{kɨ-}.
The verb also shows the complementary distribution of third person \obj{n-} and preceding objects typical of transitive verbs in Cariban \parencite[60--81]{gildea1998}.
Due to its semantics, these objects are either ideophones or direct speech \exref{hixsay}.
\input{examples/hixsay}
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}), while does not occur in \exref{hixsay.ordpl-35} where the verb is preceded by direct speech.
In the rest of the family, at least the \trio reflex shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

@@

\input{examples/sayintr}

@@

\input{examples/mivida-12}

@@

\input{examples/car-18}
%
\posscite{gildea2007greenberg} reconstruction \rc{ətepɨ} can be segmented into an optional prefix \rc{ət-} and a root \rc{epɨ}, since reflexes can be grouped into those with a reflex of \rc{ət-} and those without \pcref{tab:come}.
As seen in \cref{tab:comepara}, this division can exist within a single paradigm.

@@

\input{examples/gowhere}
%
However, a family-wide perspective reveals a more complicated story \pcref{tab:godown}.%
\footnote{
The cognacy status of parenthesized forms in \cref{tab:godown} is uncertain.
The reconstruction of \PPek \rc{ɨptə} treats the additional elements in daughter languages as non-cognate.
\textcite{meira2005southern} identify no correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}, and at least the addition of a final \rc{ŋ} in \PXin is attested elsewhere:
		\begin{inlinelist}
			\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
			\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
			\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
		\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
While a verb \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about its class.
Its reflexes in languages that preserve the split-\gl{s} system are distributed fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

The verb shows traits of both classes in \wayana, making it a \dbqu{mixed} verb synchronically.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.
\input{examples/waygodown}
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the \PC  causativizer \rc{-ka} was restricted to \gl{s_p_} verbs \parencite{gildea2019overview}.
These patterns point to \qu{to go down} being a regular \gl{s_p_} verb in pre-\wayana, but partially switching to the \gl{s_a_} class by taking a \gl{1+2}\gl{s_a_} prefix and the \gl{s_a_} class marker.
This in turn implies that (inflectionally defined) \gl{s_a_} reflexes in other languages fully switched from \gl{s_p_}.
